Finite element analysis ofmulti-dimensional and simplifiedmodels for beams and plates

### 2023-11-18 files/Chapter2/generalvibrationmodel.tex
small fixes
chapter 3

Using the bilinear forms, $b$ and $c$ from Section 1.4, Problem T-2V can be rewritten in the following form
\begin{eqnarray}
	c(\ddot{u}(t),v) + b(u(t),v) & = & (Q(t),v) \label{eq:existence:PreWV}
\end{eqnarray} for each $v \in T[0,1]$.\\

To apply the theory from \cite{VV02}, complete function spaces are required. It is known that $L^2(0,1)$ is a complete function space and hence $X$ is complete.\\

@@

The test functions $T[0,1]$ do not form a complete space. Let $V(0,1)$ be the closure of {$F_1(0,1)$} in $H^1(0,1)$ (see Section 1.4.4). It follows that $V(0,1)$ is complete and the product space {$V(0,1)\times V(0,1)$}, denoted by $V$, is complete. It is called the energy space.\\

@@

\subsubsection{Problem T-2W}\label{sssec:existence:ProblemT2W}
Find a function $u$ such that $\forall \ t \in J$, $u(t) \in V$, $\ddot{u}(t) \in W$ and
\begin{eqnarray*}
	c(\ddot{u}(t),v) + b(u(t),v) & = & (Q(t),v)
\end{eqnarray*}
for all $v\in V$. The initial conditions $u(0) = u_0$ and $\dot{u}(0)=u_1$ must be specified.\\

The space X with the inner-product $c$ is denoted by $W$. $W$ is called the inertia space and in Section 2.3 it is shown that $c$ is an inner-product for $W$.\\

@@

The following notation is important for the theory that follows.\\

Let $J$ be a interval of real numbers containing zero. It can have one of the following forms $[0,\ T)$, $[0,\ \infty)$ or an arbitrary open interval. For any function $u$ on the interval $J$ and range in a Hilbert space Z, derivatives can be defined. A derivative $u'$ for $u$ may be defined by
\begin{eqnarray*}
	||(h^{-1} u(t+h)-u(t))-u'(t)||_Z \rightarrow 0.
\end{eqnarray*} Then $u''$ is defined by $u'' = (u')'$.\\

@@

\textbf{Remark} For Problem T-2, $Z$ can be $X$, $W$ or $V$.\\

@@

\subsubsection*{Problem GVar}\label{sssec:existence:ProblemGVar}
Given a function $f:J\rightarrow X$, find a function $u\in C(J,\ X)$ such that $u'$ is continuous at $0$ with respect to $\Vert \cdot \Vert_{W}$ and for each $t\in J,\ u(t)\in V,\ u'(t) \in V,\ u''(t)\in W$ and
\begin{eqnarray}
	c(u''(t),v)+a(u'(t),v)+b(u(t),v)= (f(t),v)_{X} \ \ \ \ \textrm{for each} \ v \in V, \label{eq:existence:ProblemGVar}
\end{eqnarray}
while $u(0)=u_{0},\ \ \ u'(0)=u_{1}$.\label{sym:a} \\

@@

This theorem allows for a solution of the abstract variational problem Problem GVar, if the assumptions \textbf{A1}-\textbf{A4} is satisfied and the initial values $u_0$ and $u_1$ are admissible. It is not always easy to verify that \eqref{bilinear_equation} is satisfied.\\

@@

Semi-group theory is used in \cite{VV02} to investigate the solvability of Problem IVP and obtain a solution for Problem IVP. It is also important to mention that in \cite{VV02}, the authors provide the necessary result that shows the function $F$ is uniquely defined by $f$.\\

As mentioned before, if $a = 0$ then the inequality in \eqref{weak_damping} hold trivially. However, the operator A is defined by the operator $\Lambda$ and in the definition of $\Lambda$, the form $a$ is used (see \eqref{author_mapping}). It is proved in \cite{VV02} that equation \eqref{author_mapping} is uniquely solvable and hence $\Lambda$ is well defined. The proof remains unchanged as presented for $a$ identically zero.\\

To apply Theorem 2, the admissible initial conditions are $u_0 \in E_b$ and $u_1 \in V$. As mentioned above, this implies $\langle u_0, u_1 \rangle \in D(A)$. From semigroup theory, it follows that $U(t) \in D(A)$ for each $t$. Therefore $u(t) \in E_b$ and $u'(t) \in V$ for each $t$.\\

@@

\section{Application: Timoshenko beam model}\label{sec:existence:Application}
In this section the theory of \cite{VV02} (presented in Section \ref{sec:existence:ExistenceAndUniqueness}) is applied to Problem T-2W. This is an continuation of the example from section \ref{sec:existence}.\\

@@

To apply the theory, Problem T-2W must satisfy assumptions \textbf{A1}-\textbf{A4} and the initial values must be admissible. We must show that the assumptions are satisfied.\\

@@

Let $u \in C_0^\infty(0,1)$. Then $u(0) = u(1) = 0$ and therefore $u \in T(0,1)$ and $C_0^\infty(0,1) \subset T(0,1)$. Recall that $V(0,1)$ is the closure of $T(0,1)$ in $H^1(0,1)$. So it follows that $C_0^\infty(0,1) \subset V(0,1) \subset H^1(0,1)$. And since $C_0^\infty(0,1)$ is dense in $L_2(0,1)$, both $V(0,1)$ and $H^1(0,1)$ are dense in $L_2(0,1)$. Therefore $V(0,1)\times V(0,1)$ is dense in $X = L_2(0,1) \times L_2(0,1)$.\label{sym:Cinftyzero} \\

@@

Since $c$ is a bilinear form and by the inequality above, $c$ is an inner-product for $W$. The norm of $W$ is defined as $||\cdot||_W = \sqrt{c(\cdot,\cdot)}$. Let $C_1 = \min\left\{1, \frac{1}{\alpha}\right\}$ and $C_2 = \max\left\{1, \frac{1}{\alpha}\right\}$. Then
\begin{eqnarray}
	C_1||x||_X \leq ||x||_W \leq C_2||x||_X \label{C}
\end{eqnarray} for all $x \in X$.\\

@@

Chapter 6 is a discussion and extension of the article \cite{LVV09}. This article compares a Timoshenko beam model to a two-dimensional beam model. This is then extended to a comparison of a two-dimensional beam model to a three-dimensional beam model as well as a comparison of a two-dimensional plate model to a three-dimensional plate model.

These comparisons are made in Chapter 4 and 6. Being able to express the solutions as valid series solutions, enable us to compare the different models by only considering the eigenvalues and eigenfunctions of the different models. This is explained in detail in \cite{CVV18}.

>>>>

Using the bilinear forms, $b$ and $c$ from Section 1.4, Problem T-2V can be rewritten in the following form
\begin{eqnarray}
	c(\ddot{u}(t),v) + b(u(t),v) & = & (Q(t),v), \label{eq:existence:PreWV}
\end{eqnarray} for each $v \in T[0,1]$.

To apply the theory from \cite{VV02}, complete function spaces are required. It is known that $L^2(0,1)$ is a complete function space and hence $X$ is complete.

@@

The test functions $T[0,1]$ do not form a complete space. Let $V(0,1)$ be the closure of {$F_1(0,1)$} in $H^1(0,1)$ (see Section 1.4.4). It follows that $V(0,1)$ is complete and the product space {$V(0,1)\times V(0,1)$}, denoted by $V$, is complete. It is called the energy space.

@@

\subsubsection{Problem T-2W}\label{sssec:existence:ProblemT2W}
Find a function $u$ such that $\forall \ t \in J$, $u(t) \in V$, $\ddot{u}(t) \in W$ and
\begin{eqnarray*}
	c(\ddot{u}(t),v) + b(u(t),v) & = & (Q(t),v)
\end{eqnarray*}
for all $v\in V$. The initial conditions $u(0) = u_0$ and $\dot{u}(0)=u_1$ must be specified.

The space X with the inner-product $c$ is denoted by $W$. $W$ is called the inertia space and in Section 2.3 it is shown that $c$ is an inner-product for $W$.

@@

The following notation is important for the theory that follows.

Let $J$ be a interval of real numbers containing zero. It can have one of the following forms $[0,\ T)$, $[0,\ \infty)$ or an arbitrary open interval. For any function $u$ on the interval $J$ and range in a Hilbert space Z, derivatives can be defined. A derivative $u'$ for $u$ may be defined by
\begin{eqnarray*}
	||(h^{-1} u(t+h)-u(t))-u'(t)||_Z \rightarrow 0.
\end{eqnarray*} Then $u''$ is defined by $u'' = (u')'$.

@@

\textbf{Remark} For Problem T-2, $Z$ can be $X$, $W$ or $V$.

@@

\subsubsection*{Problem GVar}\label{sssec:existence:ProblemGVar}
Given a function $f:J\rightarrow X$, find a function $u\in C(J,\ X)$ such that $u'$ is continuous at $0$ with respect to $\Vert \cdot \Vert_{W}$ and for each $t\in J,\ u(t)\in V,\ u'(t) \in V,\ u''(t)\in W$ and
\begin{eqnarray}
	c(u''(t),v)+a(u'(t),v)+b(u(t),v)= (f(t),v)_{X} \ \ \ \ \textrm{for each} \ v \in V, \label{eq:existence:ProblemGVar}
\end{eqnarray}
while $u(0)=u_{0},\ \ \ u'(0)=u_{1}$.\label{sym:a}

@@

This theorem allows for a solution of the abstract variational problem Problem GVar, if the assumptions \textbf{A1}-\textbf{A4} is satisfied and the initial values $u_0$ and $u_1$ are admissible. It is not always easy to verify that \eqref{bilinear_equation} is satisfied.

@@

Semi-group theory is used in \cite{VV02} to investigate the solvability of Problem IVP and obtain a solution for Problem IVP. It is also important to mention that in \cite{VV02}, the authors provide the necessary result that shows the function $F$ is uniquely defined by $f$.

As mentioned before, if $a = 0$ then the inequality in \eqref{weak_damping} hold trivially. However, the operator A is defined by the operator $\Lambda$ and in the definition of $\Lambda$, the form $a$ is used (see \eqref{author_mapping}). It is proved in \cite{VV02} that equation \eqref{author_mapping} is uniquely solvable and hence $\Lambda$ is well defined. The proof remains unchanged as presented for $a$ identically zero.

To apply Theorem 2, the admissible initial conditions are $u_0 \in E_b$ and $u_1 \in V$. As mentioned above, this implies $\langle u_0, u_1 \rangle \in D(A)$. From semigroup theory, it follows that $U(t) \in D(A)$ for each $t$. Therefore $u(t) \in E_b$ and $u'(t) \in V$ for each $t$.

@@

\section{Application: Timoshenko beam model}\label{sec:existence:Application}
In this section the theory of \cite{VV02} (presented in Section \ref{sec:existence:ExistenceAndUniqueness}) is applied to Problem T-2W. This is an continuation of the example from section \ref{sec:existence}.

@@

To apply the theory, Problem T-2W must satisfy assumptions \textbf{A1}-\textbf{A4} and the initial values must be admissible. We must show that the assumptions are satisfied.

@@

Let $u \in C_0^\infty(0,1)$. Then $u(0) = u(1) = 0$ and therefore $u \in T(0,1)$ and $C_0^\infty(0,1) \subset T(0,1)$. Recall that $V(0,1)$ is the closure of $T(0,1)$ in $H^1(0,1)$. So it follows that $C_0^\infty(0,1) \subset V(0,1) \subset H^1(0,1)$. And since $C_0^\infty(0,1)$ is dense in $L_2(0,1)$, both $V(0,1)$ and $H^1(0,1)$ are dense in $L_2(0,1)$. Therefore $V(0,1)\times V(0,1)$ is dense in $X = L_2(0,1) \times L_2(0,1)$.\label{sym:Cinftyzero}

@@

Since $c$ is a bilinear form and by the inequality above, $c$ is an inner-product for $W$. The norm of $W$ is defined as $||\cdot||_W = \sqrt{c(\cdot,\cdot)}$. Let $C_1 = \min\left\{1, \frac{1}{\alpha}\right\}$ and $C_2 = \max\left\{1, \frac{1}{\alpha}\right\}$. Then
\begin{eqnarray}
	C_1||x||_X \leq ||x||_W \leq C_2||x||_X \label{C}
\end{eqnarray} for all $x \in X$.

@@

Sections 4.5 and 6.2 are a discussion and extension of the article \cite{LVV09}. This article compares a Timoshenko beam model to a two-dimensional beam model. This is then extended to a comparison of a two-dimensional beam model to a three-dimensional beam model as well as a comparison of a two-dimensional plate model to a three-dimensional plate model.

Being able to express the solutions as valid series solutions, enable us to compare the different models by only considering the eigenvalues and eigenfunctions of the different models. This is explained in detail in \cite{CVV18}.

### 2023-11-18 files/Chapter3/dynamicconvergence.tex
small fixes
chapter 3

In this chapter, two important convergence results of the Finite Element Method
(FEM) is discussed. The first result looks at the convergence of the Galerkin
approximation for second order hyperbolic type problems (or general vibation problems). The papers considered are \cite{BV13} and \cite{BSV17}. The second result is from the textbook~\cite{SF73} and examines the
convergence of eigenvalues and eigenfunctions for a one-dimensional vibration
problem, using the Finite Element Method.

@@

In Section 2.2, Problem GVar is presented and identical to the problem in
\cite{BV13}. The problem is repeated here for convenience.

\subsection{Formulation of the Galerkin approximation}
\subsubsection*{Problem GVar}
Given a function $f:J\rightarrow X$, find a function $u\in C(J,\ X)$ such that $u'$ is continuous at $0$ with respect to
$\Vert \cdot \Vert_{W}$ and for each $t\in J,\ u(t)\in V,\ u'(t) \in V,\ u''(t)\in W$ and
\begin{eqnarray}
	c(u''(t),v)+a(u'(t),v)+b(u(t),v)= (f(t),v)_{X} \ \ \ \ \textrm{for each} \ v \in V, \label{GV_PB1}
\end{eqnarray}
while $u(0)=u_{0}$, $u'(0)=u_{1}$.

@@

Recall from Chapter 2 that $V$ is the closure of the test function space $T$ in
the Hilbert space $H$.

\subsubsection*{Problem GVar$^h$}
Given a function $f: J \rightarrow X$, find a function $u_h \in C^2(J)$ such that for each $t\in J$
\begin{eqnarray}
	c(u_h''(t),v)+a(u_h'(t),v)+b(u_h(t),v)= (f(t),v)_{X} \ \ \ \ \textrm{for each} \ v \in S^h, \label{DC_E2}
\end{eqnarray}
with the initial values $u_h(0)=u^h_{0}$ and $u'_h(0)=u^h_{1}$. The initial conditions $u^h_{0}$ and $u^h_{1}$ are projections of $u_0$ and $u_1$ into the finite dimensional space $S^h$.

If the function $f$ is continuous, then Problem GVar$^h$ is equivalent to a
system of second order differential equations. The following theorem from
\cite{BV13} proves that if $f$ is continuous, then problem Problem GVar$^h$ has
a unique solution.

\subsection{System of ordinary differential equations}
Consider the standard FEM matrices defined by
\begin{eqnarray*}
	K_{ij} & = & b(\phi_j, \phi_i)\\
	C_{ij} & = & a(\phi_j, \phi_i)\\
	M_{ij} & = & c(\phi_j, \phi_i)\\
	F_{i}(t) & = & b(f(t), \phi_i)
\end{eqnarray*}

Using these matrices, Problem GVar$^h$ is rewritten as a system of ordinary
differential equations denoted by Problem ODE. \subsubsection*{Problem ODE}
Find a function $\bar{u} \in S^h$ such that \label{sym:baru}
\begin{eqnarray}
	M\bar{u}'' + C \bar{u}' + K\bar{u} = F(t) \ \ \ \textrm{ with } \bar{u}(0) = \bar{u}^h_0 \textrm{ and }  \bar{u}(1) = \bar{u}^h_1 \label{ODE}
\end{eqnarray}
The function $F$ is the interpolation of the function $f$ into the discrete space $S^h$.

@@

\subsection{Error estimates}\label{e_est}
Under the assumptions \textbf{A1}-\textbf{A4} of Section 2 and continuity of $f$, there exists unique solutions for Problem GVar and Problem GVar$^h$. The next step is to show that the solution of Problem GVar$^h$ converges to the solution of Problem GVar. This is proven the main result of \cite{BV13}.

@@

To simplify the proof, the problem is divided into a semi-discrete problem and a fully discrete problem. The two terms are investigated as seperate problems
in \cite{BV13}.

Consider the time interval $J = [0,T]$. Divide $J$ into $N$ steps with length $\tau = \frac{T}{N}$. Each interval can be expressed as $[t_{k-1}, \ t_k]$ for $k = 1,...,N$. Denote the approximation of $u_h$ on the interval $[t_{k-1}, \ t_k]$ by $u_k^h$, i.e. $u_h(t_k) = u_k^h$ for each $k$.

@@

\textbf{Fully discrete Galerkin finite element approximation}

From Theorem 2 of the semi-discrete problem an error estimate for
$[u(t_k)-u_h(t_k)]$ with respect to the norm of $W$ was obtained. The authors
of \cite{BV13} then proceed to obtain an error estimate for $[u_h(t_k) -
			u^h_k]$. Since the dimension of $S^h$ is not fixed, the equivalence of norms
cannot be used, and therefore this error estimate for $[u_h(t_k) - u^h_k]$
should also be with respect to the norm of $W$.

>>>>

In this chapter, two important convergence results of the Finite Element Method
(FEM) are discussed. The first result looks at the convergence of the Galerkin
approximation for second order hyperbolic type problems (or general vibation problems). The papers considered for the first result are \cite{BV13} and \cite{BSV17}. The second result is from the textbook~\cite{SF73} and examines the
convergence of eigenvalues and eigenfunctions for a vibration
problem, using the Finite Element Method.

@@

In Section 2.2 of this dissertation, Problem GVar is presented. It is identical to the problem in
\cite{BV13}. For convenience, the problem is repeated here.

\subsection{Formulation of the Galerkin approximation}
Recall the spaces $V$, $W$ and $X$ from Section \ref{ssec:existence:VariationlApproach} where $V \subset W \subset X$.

\subsubsection*{Problem GVar}
Given a function $f:J\rightarrow X$, find a function $u\in C(J,\ X)$ such that $u'$ is continuous at $0$ with respect to
$\Vert \cdot \Vert_{W}$ and for each $t\in J,\ u(t)\in V,\ u'(t) \in V,\ u''(t)\in W$ and
\begin{eqnarray}
	c(u''(t),v)+a(u'(t),v)+b(u(t),v)= (f(t),v)_{X} \ \ \ \ \textrm{for each} \ v \in V, \label{GV_PB1}
\end{eqnarray}
while $u(0)=u_{0}$, $u'(0)=u_{1}$.

@@

\subsubsection*{Problem GVar$^h$}
Given a function $f: J \rightarrow X$, find a function $u_h \in C^2(J)$ such that for each $t\in J$
\begin{eqnarray}
	c(u_h''(t),v)+a(u_h'(t),v)+b(u_h(t),v)= (f(t),v)_{X} \ \ \ \ \textrm{for each} \ v \in S^h, \label{DC_E2}
\end{eqnarray}
with the initial values $u_h(0)=u^h_{0}$ and $u'_h(0)=u^h_{1}$. The initial conditions $u^h_{0}$ and $u^h_{1}$ are projections of $u_0$ and $u_1$ in the finite dimensional space $S^h$.

\subsection{System of ordinary differential equations}
Problem GVar$^h$ is equivalent to a system of second order differential equations.

Consider the standard FEM matrices defined by
\begin{eqnarray*}
	K_{ij} & = & b(\phi_j, \phi_i)\\
	C_{ij} & = & a(\phi_j, \phi_i)\\
	M_{ij} & = & c(\phi_j, \phi_i)\\
	F_{i}(t) & = & b(f(t), \phi_i)
\end{eqnarray*}

Using these matrices, Problem GVar$^h$ is rewritten as a system of ordinary
differential equations denoted by Problem ODE.

Recall that $u^h(t) = \sum_{k} \bar{u}_k(t)\phi_k$.

\subsubsection*{Problem ODE}
Find a function $\bar{u} \in S^h$ such that \label{sym:baru}
\begin{eqnarray}
	M\bar{u}'' + C \bar{u}' + K\bar{u} = F(t) \ \ \ \textrm{ with } \bar{u}(0) = \bar{u}^h_0 \textrm{ and }  \bar{u}(1) = \bar{u}^h_1 \label{ODE}
\end{eqnarray}
The function $F$ is the interpolation of the function $f$ into the discrete space $S^h$.

@@

The following theorem from \cite{BV13} proves that if $f$ is continuous, then problem Problem GVar$^h$ has a unique solution.

Consider the time interval $J = [0,T]$. Divide $J$ into $N$ steps with length $\tau = \frac{T}{N}$. Each interval can be expressed as $[t_{k-1}, \ t_k]$ for $k = 1,...,N$. Denote the approximation of $u_h$ on the interval $[t_{k-1}, \ t_k]$ by $u_k^h$, i.e. $u_h(t_k) = u_k^h$ for each $k$.  A finite difference method is used to compute $u^h_k$ for each $k$.

\subsection{Error estimates}\label{e_est}
In this subsection we consider estimates for the error $u(t_k) - u^h_k$. To simplify the process, the error is divided into errors for the semi-discrete problem and the fully discrete problem.

Under the assumptions \textbf{A1}-\textbf{A4} of Section 2 and continuity of $f$, there exists a unique solution for Problem GVar$^h$. The next step is to show that the solution of Problem GVar$^h$ converges to the solution of Problem GVar.

@@

@@

From Theorem 2 of the semi-discrete problem an error estimate for
$e_k = [u(t_k)-u_h(t_k)]$ with respect to the norm of $W$ was obtained. The authors of \cite{BV13} then proceed to obtain an error estimate for $[u_h(t_k) - u^h_k]$.

Since the dimension of $S^h$ is not fixed, the equivalence of norms
cannot be used, and therefore this error estimate for $[u_h(t_k) - u^h_k]$
should also be with respect to the norm of $W$.
